Alex's Notes and Text Fragments for the JSS Manuscript

### 2021-07-12 paper/paper.Rmd
ask why feature clustering is not supported
remove references from abstract

---
documentclass: jss
author:
  - name: Patrick Schratz
    affiliation: 'Friedrich Schiller University Jena'
    # affiliaton2: 'Ludwig-Maximilians-University Munich'
    # use this syntax to add text on several lines
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{patrick.schratz@uni-jena.de}
  - name: Alexander Brenning
    affiliation: 'Friedrich Schiller University Jena \AND'
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{alexander.brenning@uni-jena.de}
  - name: Marc Becker
    affiliation: 'Ludwig-Maximilians-University Munich'
    address: |
      Department of Statistics
      Statistical Learning and Data Science group
    # To add another line, use \AND at the end of the previous one as above
  - name: Michel Lang
    affiliation: TU Dortmund University
    address: |
      | Faculty of Statistics
title:
  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal resampling methods for \\pkg{mlr3}"
  # If you use tex in the formatted title, also supply version without
  plain:     "mlr3spatiotempcv: spatiotemporal resampling methods for mlr3"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine learning framework [mlr3](https://mlr3.mlr-org.com) [@mlr3].
  It adds support for spatiotemporal resampling methods and respective visualization methods.
  Spatiotemporal resampling methods are needed to account for the existence of different autocorrelation types (spatial autocorrelation (SAC), temporal autocorrelation (TAC) or spatiotemporal autocorrelation (STAC)) in datasets.
  These autocorrelation types cause biased performance estimates of varying magnitude in cross-validation when ignored.
  At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST [@cast], blockCV [@blockCV] or sperrorest [@brenning2012].
  There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used [@roberts2017; @pohjankukka2017].

@@

In traditional cross-validation, the partitioning is based on uniform random sampling, which ignores spatial or temporal autocorrelation or any existing grouping structure as well as the structure of the prediction task, as outlined in section \@ref(introduction), and may result in over-optimistic performance estimates.
Several approaches have therefore been proposed in the literature and implemented in software to accommodate a variety of predictive situations (Table \@ref(tab:sptcv-methods)).

@@

Within the \pkg{mlr3} ecosystem, partitioning strategies are represented by their own objects of class \texttt{Resampling}, most of which are available within \pkg{mlr3} itself (e.g., regular random CV); other specialized strategies (e.g., \texttt{TaskClassifST} or \texttt{TaskRegrST}) are defined in extension packages such as \pkg{mlr3spatiotempcv}.
In the machine learning pipeline, these objects define the data splits used for model evaluation and selection (hyperparameter tuning) by machine learning algorithms.
Spatial and spatiotemporal partitioning techniques in \pkg{mlr3spatiotempcv} are currently mostly imported and interfaced from other packages, in particular \pkg{sperrorest}, \pkg{blockCV} and \pkg{CAST} [@blockCV; @brenning2012; @cast], in order to expose them to \pkg{mlr3} functionality.
To reduce first-level dependencies, some methods were re-implemented instead of importing them from upstream packages.

@@

Spatial partitioning methods can be roughly divided into three groups: (1) buffering, (2) blocking and (3) clustering methods (Table \@ref(tab:sptcv-methods)).
If a method could have been placed into multiple categories, we deliberately assigned the group which matched best the main characteristic of the respective method.

@@

<!--JM: not entirely sure what scattered appearance refers to-->
Buffering is most often applied in coordinate space due to the scattered appearance of observations in this space.
However, this method could be easily extended to a 2D + time setting.
<!-- JM: I guess you mean that you could have placed buffering also into spatiotemporal partition. I would advise to say so instead of letting the reader guess.-->
Since \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods (see Table \@ref(tab:sptcv-methods)).

@@

```{r buffer, fig.cap="Visualization of the 'Spatial buffering' method from package `blockCV`.", out.width="40%", fig.pos="ht"}
library("mlr3")
library("mlr3spatiotempcv")
task = tsk("ecuador")
resampling_buffer = rsmp("spcv_buffer", theRange = 1000)
resampling_buffer$instantiate(task)

@@

The "Spatial blocking" from package \pkg{blockCV} comes with various blocking options such as random (default) (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard pattern (`selection = "checkerboard"`).
Additionally, the user might supply a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) which will ensure a quadratic block pattern across the study area.
<!--JM: to stay consistent please add the argument which lets the user indicate a user-defined polygon or are you referring to the rectangular block design specified with rows and cols? If so, say so. If not, clarify and additionally mention the rectangular block design here-->
This kind of blocking might be of special interest for SDM studies as well as the possibility to supply a user-defined polygon with predefined blocking zones.

After the initial block creation, multiple "blocks" are combined to form a partition in correspondence with the number of folds requested by the user.
For example, in a five-fold scenario with a grid of four columns and five rows, the method would combine one fifth of all blocks (= four blocks) into a single partition.

@@

The idea of cluster analysis, which originated in the 1930s in the field of anthropology [@driver1932, @zubin1938], is to group objects with similar characteristics.
In the spatial context, the similarity of spatial coordinates is used to form such groups.
Different cluster algorithms such as hierarchical, density-based, grid-based or centroid-based clustering exist [@gower1967].

@@

The so-called "Spatial CV" method after [@russ2010; @brenning2012] and implemented in package \pkg{sperrorest} uses the coordinates of all observations to create clusters in the spatial domain with the help of the $k$-means clustering algorithm.^[$k$-means clustering belongs to the centroid-based clustering family.]
To avoid confusion with other spatial partitioning methods, we refer to this as coordinate-based clustering, or as `spcv_coords` in \pkg{mlr3spatiotempcv}, instead of using the generic "Spatial CV" term.

@@

Group-level CV follows the principle of an inherit grouping structure of the observations, both in space and/or time, to determine (groups of) partitions.
In contrast to blocking or clustering not the spatial or temporal location is used but additional metadata such as the membership to a specific crop type [@brenning2006].
The idea to follow an inherited grouping structure of the data is also referred to as "target-oriented validation" [@cast].
Even though flexibility with respect to fold creation is limited due to the inherited grouping structure within the data, some implementations allow to specify the resulting number of folds by omitting parts of the dataset.

>>>>

---
documentclass: jss
author:
  - name: Patrick Schratz
    affiliation: 'Friedrich Schiller University Jena'
    # affiliaton2: 'Ludwig-Maximilians-University Munich'
    # use this syntax to add text on several lines
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{patrick.schratz@uni-jena.de}
  - name: Alexander Brenning
    affiliation: 'Friedrich Schiller University Jena \AND'
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{alexander.brenning@uni-jena.de}
  - name: Marc Becker
    affiliation: 'Ludwig-Maximilians-University Munich'
    address: |
      Department of Statistics
      Statistical Learning and Data Science group
    # To add another line, use \AND at the end of the previous one as above
  - name: Michel Lang
    affiliation: TU Dortmund University
    address: |
      | Faculty of Statistics
title:
  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal resampling methods for \\pkg{mlr3}"
  # If you use tex in the formatted title, also supply version without
  plain:     "mlr3spatiotempcv: spatiotemporal resampling methods for mlr3"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine learning framework [mlr3](https://mlr3.mlr-org.com).
  It adds support for spatiotemporal resampling methods and respective visualization methods.
  Spatiotemporal resampling methods are needed to account for the existence of different autocorrelation types (spatial autocorrelation (SAC), temporal autocorrelation (TAC) or spatiotemporal autocorrelation (STAC)) in datasets.
  These autocorrelation types cause biased performance estimates of varying magnitude in cross-validation when ignored.
  At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST, blockCV or sperrorest.
  There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used.

@@

In traditional cross-validation, the partitioning is based on uniform random sampling, which ignores spatial or temporal autocorrelation or any existing grouping structure as well as the structure of the prediction task, as outlined in the Introduction, and may result in over-optimistic performance estimates.
Several approaches have therefore been proposed in the literature and implemented in software to accommodate a variety of predictive situations (Table \@ref(tab:sptcv-methods)).

@@

Within the \pkg{mlr3} ecosystem, partitioning strategies are represented by their own objects of class \texttt{Resampling}, most of which are available within \pkg{mlr3} itself (e.g., regular random CV); other specialized strategies (e.g., \texttt{TaskClassifST} or \texttt{TaskRegrST}) are defined in extension packages such as \pkg{mlr3spatiotempcv}.
In the machine learning pipeline, these objects define the data splits used for model evaluation and selection (hyperparameter tuning) by machine learning algorithms.
Spatial and spatiotemporal partitioning techniques in \pkg{mlr3spatiotempcv} are currently mostly imported and interfaced from other packages, in particular \pkg{sperrorest}, \pkg{blockCV} and \pkg{CAST} [@blockCV; @brenning2012; @cast], in order to expose them to \pkg{mlr3} functionality.
To reduce first-level dependencies, some methods were re-implemented instead of imported from upstream packages.

@@

Judging by their common characteristics and the available implementations, spatial partitioning methods can be coarsely categorized into (1) buffering, (2) blocking and (3) clustering methods.
If a method had the potential to belong to multiple categories, a deliberate category assignment was made for one specific group which matched the main characteristic of the respective method best.

@@

Buffering is mostly applied in the coordinate space due to the scattered appearance of observations in this space.
However, there is no practical limitation that would prevent this method to be applied in a 2D + time setting.
Because \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods.

@@

```{r buffer, fig.cap="Visualization of the 'Spatial buffering' method from package `blockCV`.", out.width="40%", fig.pos="ht"}
task = tsk("ecuador")
resampling_buffer = rsmp("spcv_buffer", theRange = 1000)
resampling_buffer$instantiate(task)

@@

The "Spatial blocking" from package \pkg{blockCV} comes with various options to customize the zone creation.
Besides the choice of relying on an initial random  (default) (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard pattern (`selection = "checkerboard"`) for the final partition aggregation, support for SDM specific background data is built-in.
Supplying a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) will create a quadratic block pattern across the study area.
Ultimately, users can also supply a user-defined polygon which defines the zones.

After the initial block creation, multiple "blocks" are combined to form a partition, honoring the number of folds requested by the user.
For example, in a five-fold scenario with a grid of four columns and five rows, the method would combine one fifth of all blocks (= four blocks) into a single partition.

@@

The idea of cluster analysis, which originated in the 1930s in the anthropology field [@driver1932, @zubin1938], is to group objects with similar characteristics.
In the spatial context, the similarity of spatial coordinates is used to form such groups.
Different cluster algorithms such as hierarchical, density-based, grid-based or centroid-based clustering exist [@gower1967].

@@

The so-called "Spatial CV" method after [@russ2010; @brenning2012] and implemented in package \pkg{sperrorest} uses the coordinates of all observations to create clusters in the spatial domain.
$k$-means clustering, which belongs to the centroid-based clustering family, has been proposed as a clustering method.
To avoid confusion with other spatial partitioning methods, we refer to this as coordinate-based clustering, or as `spcv_coords` in \pkg{mlr3spatiotempcv}, instead of using the generic "Spatial CV" term.

@@

Group-level CV follows the principle of an inherit grouping structure of the observations, both in space and/or time, to determine (groups of) partitions.
In contrast to blocking or clustering not the the spatial or temporal location is used but additional metadata such as the membership to a specific crop type [@brenning2006].
The idea to follow an inherited grouping structure of the data is also referred to as "target-oriented validation" [@cast].
Even though flexibility with respect to fold creation is limited due to the inherited grouping structure within the data, some implementations allow to specify the resulting number of folds by omitting parts of the dataset.
